Basic Parameter estimation of Binary Neutron Star Systems by the Advanced LIGO/Virgo Network

### 2013-06-28 doc/paper.tex
Edits and comments from Will.

Within the next five years, the Advanced LIGO/Virgo network will have
reached a sensitivity sufficient to enable the routine detection of
gravitational waves.  Beyond the initial detection, the scientific
promise of these instruments relies on the ability to perform
parameter estimation. The majority of this effort has been towards the
detection and characterization of gravitational waves from compact
binary coalescence, e.g. the coalescence of binary neutron stars.
While several previous studies have focused on the parameter
estimation abilities of advanced detectors, the majority have relied
on approximation techniques such as the Fisher Matrix.  We report the
statistical uncertainties that will be achievable, in practice, using
the full parameter estimation machinery developed by the LIGO/Virgo
Collaboration.  We find the recovery of the individual masses to be
fractionally within 10\% at the 65\% credible intervals for equal-mass
systems, and within 2\% for unequal-mass systems.  We also report the
average uncertainties on the sky-locations, luminosity distance, and
orbital inclinations of randomly distributed sources that can be
achieved by different network configurations.
\end{abstract}

@@

Within the next few years, the first generation of gravitational-wave
detectors capable of regularly resolving astrophysical sources will
come online \citep{AdvLIGO,AdvVirgo}.  The Advanced LIGO and Virgo
detectors will provide the first insights into the final moment of
binary compact object mergers, including the merger of binary neutron
star systems.  Intense preparations are underway to characterize and
extract as much physical information as possible from these signals.

The mergers of binary neutron stars (BNS) are expected to be the most
common compact binary source in the advanced detector era.  Models
from stellar evolution and observations of binary pulsars suggest that
the number of binary neutron star mergers within Advanced LIGO/Virgo's
detection horizon could reach into the hundreds each year, with an
expected mean of $\sim 40$ per year \citep{RatesPaper}.  Although the
peak sensitivity of ground-based detectors is not focused on the
frequency at which BNS systems merge, it could still be possible to
extract information about both strong field gravitational physics
\carl{[CITE]} and the hydrodynamics of dense matter (e.g. the equation
of state of nuclear matter) \carl{[CITE]}.  Furthermore, the
observations of multiple BNS systems will provide key insight into the
evolution of binary systems in the field \citep{VickyRates,KimRates}
. As such, BNS systems will likely form the ``bread and butter'' of
the compact binary coalescence detection effort in the coming years.

Of course one must distinguish between the detection of such events,
and the precision measurement of the relevant physical parameters.
The detection of BNS systems will be performed with a matched
filtering approach.  By comparing the data stream with a bank of
theoretical templates, the time-series data can be searched for
candidate signals at a sufficiently rapid rate to analyze months of
data.  When the overlap between one of the template waveforms and the
detector output is sufficiently large (97\%), a detection candidate is
found.  However, the parameter space of these signals can be highly
degenerate, with several locations in parameter space corresponding to
nearly identical waveforms.  In order to fully realize the science
potential of these instruments, we must perform a full exploration of
the parameter space for each detection.  By analyzing the parameter
space of each candidate with a Bayesian inference technique, we will
be able to make precise, scientifically meaningful statements about
the physics of these systems.  To that end, we employ a Markov-Chain
Monte Carlo sampling algorithm, \texttt{lalinference\_mcmc}, included
in the LIGO Algorithm Library parameter estimation code,
\texttt{LALInference}, to analyze the parameter space of BNS systems.

By employing the full parameter-estimation machinery that will
eventually be used in the Advanced LIGO/Virgo era, our results give
the first estimates at the realistic capabilities of advanced
ground-based detectors to characterize BNS systems.  Until recently,
the majority of studies have employed the Fisher matrix formalism
which was first adapted for gravitational-wave parameter estimation by
\citep{FinnDetection}.  While each of these studies
\citep{PoissonWill,CutlerFlanagan,ArunPE} have pointed out the
limitations and flaws of the Fisher Information Matrix, there has been
no study which investigates the BNS parameter estimation capabilities
of Advanced LIGO/Virgo using the techniques that will eventually be
employed.  The recent work of \cite{Vallisneri} and our own work
\citep{Inadequacies} has demonstrated that the Fisher matrix cannot
even be treated as a lower bound on the standard deviations of certain
parameters (particularly the masses) measurable in gravitational-wave
detectors.

In this paper, we perform a systematic study of the statistical
uncertainties with which the Advanced LIGO/Virgo network will be able
to measure the basic parameters of BNS mergers.  In Section
\ref{PEsection}, we describe the machinery of our parameter estimation
code, \texttt{LALInference}, and the associated MCMC sampler,
\texttt{lalinference\_mcmc}, as well as the frequency-domain
gravitational-wave template we employ.  In Section
\ref{resultsSection}, we qualitatively analyze the posterior
probability density functions for BNS systems with different masses
and extrinsic parameters.  The results are divided into three
parameter sets of interest: the recovery of the mass parameters
(Section \ref{massSection}), the recovery of the orbital inclination
and luminosity distance (Section \ref{idSection}), and the
localization of sources on the sky (Section \ref{skySection}).
Finally we provide quantitative 1-dimensional confidence intervals on
each parameter in Section \ref{ciSection}.  These results, contained
in Tables \ref{ciTableIntrinsic} and \ref{ciTableExtrinsic}, are
intended as a reference for the optimal BNS parameter estimation
capabilities of Advanced LIGO/Virgo.  We assume that $G=c=1$
throughout.

@@

We begin by introducing a Bayesian formalism for parameter estimation.
%Before offering a brief derivation of the FIM, a few preliminary
%techniques must be introduced.  First,
We assume that the time-domain signal in a gravitational-wave network
can be written as a combination of a gravitational waveform $h_0$ and
the noise of the detector $n$.  We further assume that this noise is
stationary and Gaussian with zero mean.  Therefore, the detector
output is simply
\begin{equation}
s = n + h_0 .
\label{SignalAddition}
\end{equation}
Since the noise model is Gaussian, we can write the probability of a
specific signal realization $s$ as proportional to the probability
that the residual is Gaussian distributed once the waveform has been
subtracted
\begin{align}
  p(s | \thpara) &\propto \exp\left[-\frac{1}{2}\left<n|n
    \right>\right] \nonumber \\ &= \exp\left[-\frac{1}{2}\left < s -
    h(\thpara) | s-h(\thpara)\right >\right] ,
  \label{likelihood}
\end{align}
where $\thpara$ is the set of parameters for the template waveforms,
and $p(s | \thpara)$ is the likelihood of the signal $s$ given the
parameters $\thpara$.  The inner product, $\left< ~|~ \right> $, is
defined using the noise spectrum of the detectors as
\begin{equation}
  \left<a|b\right> \equiv 4 \Re \int
  \frac{\tilde{a}(f)\tilde{b}^*(f)}{S_n(f)} df ,
  \label{innerProduct}
\end{equation}
where $S_n(f)$ is the one-sided power spectral density (PSD) as a
function of frequency, and $\tilde{a}(f)$ and $\tilde{b}(f)$ are the
Fourier transforms of the time-domain data $a(t)$ and $b(t)$.  If we
pick a set of parameters $\thpara$ such that $h(\thpara) = h_0$, then
the posterior \eqref{posterior} will be near a global maximum;
however, our choice of prior the presence of noise will in general
deflect the maximum of our posterior away from the value at
$\thpara_0$.  Therefore, the maximum posterior parameters do not
necessarily correspond to the true parameters of the source, so when
interpreting parameter estimation results we must consider the full
posterior distribution function.

@@

We are interested in the posterior $p(\thpara | s)$ as it encodes
information about the prior state of knowledge about the problem in
addition to the likelihood of the signal.  For instance, since it is
assumed that compact binaries form homogeneously in the local universe
observable by advanced detectors, we can safely assume that the
population of sources will be distributed homogeneously in volume, or
evenly in $D^2$.  Our prior information can come from mathematical
limits in the parameter space or from our \textit{a priori} knowledge
of astrophysical systems.  We employ priors which are

\begin{itemize}
\item uniform in component masses from $0.8M_{\odot} \leq M_{1,2} \leq
  30M_{\odot}$, with a minimum chirp mass \eqref{eqRatioCM} of 0.6
  $M_{\odot}$,
\item uniform in volume, which implies a luminosity distance prior of
  $p(D) = D^2 dD$,
\item uniform in coalescence time over the segment of data being
  analyzed, and
\item uniform in all other parameters, with angular parameters being
  flat over their corresponding domains.
\end{itemize}

\noindent The mass priors listed here are the generic priors used in
follow-up analysis, not specifically for BNS systems.  While we could
have used a prior with a lower component mass cutoff for the present
study, in practice the higher prior boundary does not effect our
current analysis.

@@

We use a frequency-domain waveform accurate up to $3.5$ post-Newtonian
(pN) order in phase and 3 pN order in amplitude.  We restrict
ourselves to quasi-circular, non-spinning waveforms as a simplifying
assumption.  The standard form of our waveform model, known as the
\textit{TaylorF2} approximant, is calculated via the stationary-phase
approximation.  In this setup, the gravitational-wave amplitude is
given by
\begin{equation}
\tilde{h}(f) = a(t_f) e^{i \psi(f)},
\label{amplitude}
\end{equation}
where $a(t_f)$ is the amplitude evaluated at a stationary-phase
reference point, which to lowest order takes the form $a(t_f) \propto
f^{-7/6} \chmass^{5/6}\Theta(\text{angle})/D$, where $D$ is the
luminosity distance of the binary, and $\psi(f)$ is the pN phase.
$\Theta(\text{angle})$ is a function of the orbital orientation with
respect to the detector network in terms of the sky position, orbital
inclination, and the wave polarization.  In addition to the total
mass, $M_{tot}\equiv M_1+M_2$, it is convenient to work with the mass
ratio and chirp mass, defined by
\begin{equation}
  q\equiv M_1/M_2~~~~\text{and}~~~~\chmass = (M_1 M_2)^{3/5}
  M_{tot}^{-1/5},
  \label{eqRatioCM}
\end{equation}
respectively.\footnote{Most gravitational-wave literature instead uses
  the \textit{symmetric mass ratio}, defined as $\eta \equiv M_1 M_2 /
  M^2$.  We elect to use $q$ as it is more physically intuitive, and
  the prior $p(q)$ is uniform when the component mass priors are
  uniform.}  By convention, $M_1 \geq M_2$.  The stationary phase then
becomes an expansion in the Newtonian orbital velocity, $v=(\pi
M_{tot} f)^{1/3}$,
\begin{equation}
\psi(f) = 2 \pi f t_c - \phi_0 + \frac{\pi}{4} +
\frac{3}{128}\left(\frac{M_{tot}}{\chmass}\right)^{5/3}\sum^{n}_{k=0}\alpha_{k}v^{k-5}
\label{phase}
\end{equation}

@@

For this study, we consider four seperate populations of BNS systems,
with component mass combinations of $1M_{\odot}/1M_{\odot}$,
$1.4M_{\odot}/1.4M_{\odot}$, $1M_{\odot}/2.5M_{\odot}$, and
$2.5M_{\odot}/2.5M_{\odot}$.  Each population consisted of 40 signals
distributed randomly in sky location, polarization, inclination,
time-of-arrival, and coalescence phase.  The luminosity distance, $D$,
was selected to provide a network SNR of $\rho_{\text{network}}=20$
for each source.  While unphysical from an astrophysics perspective
(binary sources should be distributed evenly in comoving volume), this
choice allows us to explore the parameter estimation in the context of
optimal detection candidates.

Finally, note that the power-spectral density that defines the inner
product \eqref{innerProduct} is simply the time-averaged sensitivity
of a given detector to a specific frequency.  Ignoring non-Gaussian
glitches, any stretch of data should contain a specific noise
realization drawn from a Gaussian colored by the PSD.
%   However, what we are interested in is not the parameter estimation that can occur in a specific draw
  % from this probability of noises, but the uncertainties averaged
over all possible noise realizations. Without a realistic model
including the actual instrument sensitivity, non stationary noise, and
uncertainty in the PSD we cannot accurately simulate GW data to the
point of making precise predictions about parameter estimation
capabilities.  Furthermore, the posterior distribution from a future
detection will be influenced by the realization of the noise at the
time of the detection.  Because of this, we elect to focus only on the
statistical errors due to the flexibility of the waveforms and the
expected sensitivity of the advanced era detectors and suppress any
effects of detector noise by not including a noise realization in our
simulated data.  Consequently, what we recover with the MCMC could be
interpreted as the posterior distribution function for each source
averaged over all possible (Gaussian) noise realizations assuming we
have the correct PSD.  \footnote{Similarly to how the Fisher matrix
  formalism can provide the Cramer-Rao bound averaged over all noise
  realizations.  Unlike a Fisher matrix study, we make no assumptions
  about the shape of the posterior distribution function, nor the
  location of the peak, while still requiring a feasible number of
  MCMC runs to sufficiently cover parameter space, which would not be
  the case had we needed to average several noise-dependent results.}
To check that our reported statistical errors are not in gross
disagreement with those obtained in the presence of noise, we selected
a particular simulated signal to perform parameter estimation in
several noise realizations.  The comparison of the noise-dependent and
noise-free results can be found in \ref{noiseSection}.

@@

\begin{figure*}[t!]
\centering \includegraphics[trim=7cm 0cm 0cm 0cm,
  clip=true,scale=0.55]{9dpdf.pdf}
%\begin{array}{ccc} 
%\includegraphics[width=2.2in]{1dpdfs/mc.pdf} &\includegraphics[width=2.2in]{1dpdfs/eta.pdf}  & \includegraphics[width=2.2in]{1dpdfs/m1.pdf}\\
%\includegraphics[width=2.2in]{1dpdfs/time.pdf} &\includegraphics[width=2.2in]{1dpdfs/dist.pdf}  &\includegraphics[width=2.2in]{1dpdfs/iota.pdf}\\
%\includegraphics[width=2.2in]{1dpdfs/ra.pdf} & \includegraphics[width=2.2in]{1dpdfs/dec.pdf} & \includegraphics[width=2.2in]{1dpdfs/m2.pdf}
%\end{array}$
\caption{Marginalized 1D posterior probability density functions taken
  from a typical $1.4M_{\odot}/1.4M_{\odot}$ BNS system.  We have
  plotted each of the 9 parameters for our non-spinning BNS problem as
  stated in section \ref{waveformSection}.  Note how, even for
  parameters with excellent recovery, the peak of the 1D Gaussian is
  displaced from the injected value (in dashed red).  This is not due
  to a systematic bias, but is caused by the marginalization of a
  single dimension from the full 9D posterior space.  To better see
  this effect, compare the 1D PDF for $\mathcal{M}_{c}$ and $q$ with
  the 2D marginalized PDF in Figure~\ref{1414masses}}
\end{figure*}

Of the nine parameters in the domain of the waveform, only six are
particularly physically interesting: the masses of the two binaries,
$M_1$ and $M_2$, the orbital inclination, $\iota$, the angular
position on the sky, $\alpha$ and $\delta$, and the luminosity
distance of the source, $D$.  While the coalescence phase $\phi_c$,
the coalescence time $t_c$, and the wave polarization $\psi$ must be
included in any parameter estimation of the waveform, they do not
encode any information of particular astrophysical
interest.\footnote{For systems whose components are non-spinning.  The
  presence of intrinsic angular momentum will, in general, couple to
  the coalescence phase and wave polarization.}

In Figure~\ref{9dPDF}, we provide an example of the nine,
1-dimensional marginalized posterior probability density functions
recovered from a single $1.4M_{\odot}/1.4M_{\odot}$ BNS system.  These
PDFs are typical of the type of results that will be produced by
parameter estimation studies in the advanced detector era.  Notice
that the peak of several parameters, including the chirp mass,
$\chmass$, appears to be displaced from the true values in red.  This
effect is due to the reduction of the 9-dimensional PDF to a series of
marginalized 1-dimensional PDFs.  For instance, the 1-dimensional PDF
for chirp mass is marginalized via

\begin{equation}
p(\chmass | s) = \int _{\thpara \setminus \chmass} p(\thpara |
s)~d(\thpara \setminus \chmass)
\label{mcMarginalization}
\end{equation}

\noindent where the notation $\thpara \setminus \chmass$ implies all
parameter of \eqref{parameterspace} except $\chmass$.  Other parameter
and higher-dimensional marginalizations follow a similar convention.
In practice, the MCMC samples make this integral trivial: since the
samples are distributed according to the posterior,
\eqref{mcMarginalization} can be calculated by simply reading off the
relevant samples in a single parameter, implicitly calculating a
Monte-Carlo integral over all other parameters.

@@

Of the nine variables in our parameter space \eqref{parameterspace},
the two of immediate physical interest are the two parameters,
$\chmass$ and $q$, or correspondingly the direct masses of the two
companions, $M_1$ and $M_2$.  As stated above, the ability of Advanced
LIGO/Virgo to construct a population of BNS masses will be one of the
more useful applications of gravitational-wave astronomy.

We find that there is virtually no difference between the mass PDFs of
our different injected signals within each mass bin.  While this may
initially seem disconcerting, it is to be expected.  Recall that all
of the BNS signals were injected with a network SNR of 20 into a
zero-noise detector realization.  Furthermore, note that the mass
parameters are the only two which directly effect the phase of the
TaylorF2 waveform \eqref{phase}.  Therefore, as long as the injected
mass parameters are identical within each mass system, and as long as
the injected SNR is identical, the amount of recoverable information
in each mass parameter should be almost identical, and for a
sufficiently converged MCMC chain, the recovered posterior PDF should
be identical.

>>>>

Within the next five years, the Advanced LIGO/Virgo network will have
reached a sensitivity sufficient to enable the routine detection of
gravitational waves.  Beyond the initial detection, the scientific
promise of these instruments relies on the ability to perform
parameter estimation. The majority of this effort has been towards the
detection and characterization of gravitational waves from compact
binary coalescence, e.g. the coalescence of binary neutron stars.
While several previous studies have focused on the parameter
estimation abilities of advanced detectors, the majority have relied
on approximation techniques such as the Fisher Matrix.  We report the
statistical uncertainties that will be achievable, in practice, using
the full parameter estimation machinery developed by the LIGO/Virgo
Collaboration.  We find the recovery of the individual masses to be
fractionally within 10\% at the 65\% credible intervals for equal-mass
systems, and within 2\% for unequal-mass systems.  We also report the
average uncertainties on the sky-locations, luminosity distance, and
orbital inclinations of randomly distributed sources that can be
achieved by different network configurations.  \will{I'm not sure that
  we want to say this exactly this way---the sources in this study are
  not randomly distributed in SNR, but rather cluster at rather loud
  signals.}
\end{abstract}

@@

Within the next few years, the first generation of gravitational-wave
detectors capable of regularly resolving astrophysical sources will
come online \citep{AdvLIGO,AdvVirgo}.  The Advanced LIGO and Virgo
detectors will provide the first insights into the final moments of
binary compact object mergers, including the merger of binary neutron
star systems.  Intense preparations are underway to characterize and
extract as much physical information as possible from these signals.

The mergers of binary neutron stars (BNS) are expected to be the most
common compact binary source in the advanced detector era.  \will{This
  is technically true (thanks to the word ``expected''), but the
  uncertainty in rates is so large that it would not be surprising at
  all for it to be wrong.  Re-word or eliminate.} Models from stellar
evolution and observations of binary pulsars suggest that the number
of binary neutron star mergers within Advanced LIGO/Virgo's detection
horizon could reach \will{several tens to hundreds} each year
\citep{RatesPaper}.  Although the peak sensitivity of ground-based
detectors is not focused on the frequency at which BNS systems merge,
it could still be possible to extract information about both strong
field gravitational physics \carl{[CITE]} and the hydrodynamics of
dense matter (e.g. the equation of state of nuclear matter)
\carl{[CITE]}.  Furthermore, the observations of multiple BNS systems
will provide key insight into the evolution of binary systems in the
field \citep{VickyRates,KimRates} . As such, BNS systems will likely
form the ``bread and butter'' of the compact binary coalescence
detection effort in the coming years.

Of course one must distinguish between the detection of such events,
and the precision measurement of the relevant physical parameters.
The detection of BNS systems will be performed with a matched
filtering approach.  By comparing the data stream with a bank of
theoretical templates, the time-series data can be searched for
candidate signals at a sufficiently rapid rate to analyze months of
data.  When the overlap between one of the template waveforms and the
detector output is sufficiently large (97\%), a detection candidate is
found.  \will{This is not true.  The \emph{templates} are chosen so
  that any signal should have 97\% overlap with at least one template,
  but detection depends on the intrinsic strength of the signal, not
  its degree of overlap with templates.}  However, the parameter space
of these signals can be highly degenerate, with several locations in
parameter space corresponding to nearly identical waveforms.  In order
to fully realize the science potential of these instruments, we must
perform a full exploration of the parameter space for each detection.
By analyzing the parameter space of each candidate with a Bayesian
inference technique, we will be able to make precise, scientifically
meaningful statements about the physics of these systems.  To that
end, we employ a Markov-Chain Monte Carlo sampling algorithm,
\texttt{lalinference\_mcmc}, included in the LIGO Algorithm Library
parameter estimation code, \texttt{LALInference}, to analyze the
parameter space of BNS systems.

By employing the full parameter-estimation machinery that will
eventually be used in the Advanced LIGO/Virgo era, our results give
\will{\sout{the first}} estimates at the realistic capabilities of
advanced ground-based detectors to characterize BNS systems.  Until
recently, the majority of studies have employed the Fisher matrix
formalism which was first adapted for gravitational-wave parameter
estimation by \citet{FinnDetection}.  While each of these studies
\citep{PoissonWill,CutlerFlanagan,ArunPE} have pointed out the
limitations and flaws of the Fisher Information Matrix, there has been
no study which investigates the BNS parameter estimation capabilities
of Advanced LIGO/Virgo using the techniques that will eventually be
employed.  The recent work of \cite{Vallisneri} and our own work
\citep{Inadequacies} has demonstrated that the Fisher matrix cannot
even be treated as a lower bound on the standard deviations of certain
parameters (particularly the masses) measurable in gravitational-wave
detectors.

In this paper, we perform a systematic study of the statistical
uncertainties with which the Advanced LIGO/Virgo network will be able
to measure the basic parameters of BNS mergers.  In Section
\ref{PEsection}, we describe the machinery of our parameter estimation
code, \texttt{LALInference}, and the associated MCMC sampler,
\texttt{lalinference\_mcmc}, as well as the frequency-domain
gravitational-wave template we employ.  In Section
\ref{resultsSection}, we qualitatively analyze the posterior
probability density functions for BNS systems with different masses
and extrinsic parameters.  \will{Here would be a good place to have a
  sentence or two about the distribution of source parameters---the
  choices of masses, angles, and distances.}  The results are divided
into three parameter sets of interest: the recovery of the mass
parameters (Section \ref{massSection}), the recovery of the orbital
inclination and luminosity distance (Section \ref{idSection}), and the
localization of sources on the sky (Section \ref{skySection}).
Finally we provide quantitative 1-dimensional confidence intervals on
each parameter in Section \ref{ciSection}.  These results, contained
in Tables \ref{ciTableIntrinsic} and \ref{ciTableExtrinsic}, are
intended as a reference for the optimal BNS parameter estimation
capabilities of Advanced LIGO/Virgo.  We assume that $G=c=1$
throughout.

@@

We begin by introducing a Bayesian formalism for parameter estimation.
%Before offering a brief derivation of the FIM, a few preliminary
%techniques must be introduced.  First,
We assume that the time-domain signal in a gravitational-wave network
can be written as a combination of a gravitational waveform $h_0$ and
the noise of the detector $n$.  We further assume that this noise is
stationary and Gaussian with zero mean.  The detector output is
\begin{equation}
s = n + h_0 .
\label{SignalAddition}
\end{equation}
Since the noise model is Gaussian, we can write the probability of a
specific signal realization $s$ given an input waveform as
proportional to the probability that the residual is Gaussian
distributed once the waveform has been subtracted
\begin{align}
  p(s | \thpara) &\propto \exp\left[-\frac{1}{2}\left<n|n
    \right>\right] \nonumber \\ &= \exp\left[-\frac{1}{2}\left < s -
    h(\thpara) | s-h(\thpara)\right >\right] ,
  \label{likelihood}
\end{align}
where $\thpara$ is the set of parameters for the template waveforms.
The quantity $p(s | \thpara)$ is the likelihood of the signal $s$
given the parameters $\thpara$.  The inner product, $\left< ~|~
\right> $, is defined using the noise spectrum of the detectors as
\begin{equation}
  \left<a|b\right> \equiv 4 \Re \int
  \frac{\tilde{a}(f)\tilde{b}^*(f)}{S_n(f)} df ,
  \label{innerProduct}
\end{equation}
where $S_n(f)$ is the one-sided power spectral density (PSD) as a
function of frequency, and $\tilde{a}(f)$ and $\tilde{b}(f)$ are the
Fourier transforms of the time-domain data $a(t)$ and $b(t)$.  If we
pick a set of parameters $\thpara$ such that $h(\thpara) = h_0$, then
the posterior \eqref{posterior} will be near a global maximum;
however, the presence of noise will in general deflect the maximum of
our posterior away from the value at $\thpara_0$.  Therefore, the
maximum posterior parameters do not necessarily correspond to the true
parameters of the source; when interpreting parameter estimation
results it is important to consider the full posterior distribution
function.

@@

We are interested in the posterior $p(\thpara | s)$ as it encodes
information about the prior state of knowledge about the problem in
addition to the likelihood of the signal.  For instance, since it is
assumed that compact binaries form homogeneously in the local universe
observable by advanced detectors, we can safely assume that the
population of sources will be distributed homogeneously in volume, or
with a density proportional to $D^2$.  Our prior information can come
from mathematical limits in the parameter space or from our \textit{a
  priori} knowledge of astrophysical systems.  We employ priors which
are
\begin{itemize}
\item uniform in component masses from $0.8M_{\odot} \leq M_{1,2} \leq
  30M_{\odot}$, with a minimum chirp mass (see Eq.~\eqref{eqRatioCM})
  of 0.6 $M_{\odot}$,
\item uniform in volume, which implies a luminosity distance prior of
  $p(D)dD = D^2 dD$,
\item uniform in coalescence time over the segment of data being
  analyzed, and
\item a delta function at zero spin (i.e.\ we use waveforms without
  spin)
\item isotropic in all angles
\end{itemize}
The mass priors listed here are the generic priors used in follow-up
analysis, not specifically for BNS systems.  While we could have used
a prior with a lower component mass cutoff for the present study, in
practice the higher prior boundary does not effect our current
analysis.

@@

We use a frequency-domain waveform accurate up to $3.5$ post-Newtonian
(pN) order in phase and 3 pN order in amplitude.  We restrict
ourselves to quasi-circular, non-spinning waveforms as a simplifying
assumption.  The standard form of our waveform model, known as the
\textit{TaylorF2} approximant, is calculated via the stationary-phase
approximation.  In this setup, the gravitational-wave amplitude is
given by
\begin{equation}
\tilde{h}(f) = a(t_f) e^{i \psi(f)},
\label{amplitude}
\end{equation}
where $a(t_f)$ is the amplitude evaluated at a stationary-phase
reference point, which to lowest order takes the form $a(t_f) \propto
f^{-7/6} \chmass^{5/6}\Theta(\text{angle})/D$, where $D$ is the
luminosity distance of the binary, and $\psi(f)$ is the pN phase.
$\Theta(\text{angle})$ is a function of the orbital orientation with
respect to the detector network in terms of the sky position, orbital
inclination, and the wave polarization.  In addition to the total
mass, $M_{tot}\equiv M_1+M_2$, it is convenient to work with the mass
ratio and chirp mass, defined by
\begin{equation}
  q\equiv M_1/M_2~~~~\text{and}~~~~\chmass = (M_1 M_2)^{3/5}
  M_{tot}^{-1/5},
  \label{eqRatioCM}
\end{equation}
respectively.\footnote{Most gravitational-wave literature instead uses
  the \textit{symmetric mass ratio}, defined as $\eta \equiv M_1 M_2 /
  M^2$.  We elect to use $q$ as it is more physically intuitive, and
  the prior $p(q)$ is uniform when the component mass priors are
  uniform.  \will{This is not true.  I calculate (could be a bug, so
    check it)
    \begin{equation}
      p(\chmass, q) = \frac{\chmass}{q} \left(\frac{\left(1+q\right)^2}{q}\right)^{\frac{1}{5}}
    \end{equation}
    when 
    \begin{equation}
      p\left(m_1, m_2\right) \propto \mathrm{const}
    \end{equation}
    The reasons to use $q$ instead of $\eta$
    \begin{itemize}
      \item It is more intuitive to astronomers (you mention this
        above).
      \item As $m_1 \to m_2$, or $q \to 1$, or $\eta \to 1/4$, the
        prior on $\eta$ for equal mass systems diverges (it is an
        integrable singularity, but still singular), while the prior
        on $q$ is well-behaved.  This doesn't matter for sampling
        purposes, since the divergence in integrable, but it is
        disconcerting for near-equal-mass systems to see a large peak
        in the histogram for $\eta$ near $\eta = 1/4$.
    \end{itemize}}}  By convention, $M_1 \geq M_2$.  The stationary phase then
becomes an expansion in the Newtonian orbital velocity, $v=(\pi
M_{tot} f)^{1/3}$,
\begin{equation}
\psi(f) = 2 \pi f t_c - \phi_0 + \frac{\pi}{4} +
\frac{3}{128}\left(\frac{M_{tot}}{\chmass}\right)^{5/3}\sum^{n}_{k=0}\alpha_{k}v^{k-5}
\label{phase}
\end{equation}

@@

For this study, we consider four seperate populations of BNS systems,
with component mass combinations of $1M_{\odot}/1M_{\odot}$,
$1.4M_{\odot}/1.4M_{\odot}$, $1M_{\odot}/2.5M_{\odot}$, and
$2.5M_{\odot}/2.5M_{\odot}$.  Each population consisted of 40 signals
distributed randomly in sky location, polarization, inclination,
time-of-arrival, and coalescence phase.  The luminosity distance, $D$,
was selected to provide a network SNR of $\rho_{\text{network}}=20$
for each source.  While unphysical from an astrophysics perspective
(binary sources should be distributed evenly in comoving volume), this
choice allows us to explore the parameter estimation in the context of
\will{\sout{optimal}} \will{loud, but plausible} detection candidates.

Finally, note that the power-spectral density that defines the inner
product \eqref{innerProduct} is simply the time-averaged sensitivity
of a given detector to a specific frequency.  Ignoring non-Gaussian
glitches, any stretch of data should contain a specific noise
realization drawn from a Gaussian colored by the PSD.  Without a
realistic model including the actual instrument sensitivity, non
stationary noise, and uncertainty in the PSD we cannot accurately
simulate GW data to the point of making precise predictions about
parameter estimation capabilities.  Furthermore, the posterior
distribution from a future detection will be influenced by the
realization of the noise at the time of the detection.  Because of
this, we elect to focus only on the statistical errors due to the
flexibility of the waveforms and the expected sensitivity of the
advanced era detectors and suppress any effects of detector noise by
picking the (most likely) noise realization that has $n(t) \equiv 0$.
Consequently, what we recover with the MCMC could be interpreted as
the posterior distribution function for each source averaged over all
possible (Gaussian) noise realizations assuming we have the correct
PSD.  \will{This is not true because the map from noise realization to
  posterior is not linear.}

To check that our reported statistical errors are not in gross
disagreement with those obtained in the presence of noise, we selected
a particular simulated signal to perform parameter estimation in
several noise realizations.  The comparison of the noise-dependent and
noise-free results can be found in \ref{noiseSection}.

@@

\begin{figure*}[t!]
\centering \includegraphics[trim=7cm 0cm 0cm 0cm,
  clip=true,scale=0.55]{9dpdf}
%\begin{array}{ccc} 
%\includegraphics[width=2.2in]{1dpdfs/mc.pdf} &\includegraphics[width=2.2in]{1dpdfs/eta.pdf}  & \includegraphics[width=2.2in]{1dpdfs/m1.pdf}\\
%\includegraphics[width=2.2in]{1dpdfs/time.pdf} &\includegraphics[width=2.2in]{1dpdfs/dist.pdf}  &\includegraphics[width=2.2in]{1dpdfs/iota.pdf}\\
%\includegraphics[width=2.2in]{1dpdfs/ra.pdf} & \includegraphics[width=2.2in]{1dpdfs/dec.pdf} & \includegraphics[width=2.2in]{1dpdfs/m2.pdf}
%\end{array}$
\caption{\label{9dPDF} Marginalized 1D posterior probability density
  functions taken from a typical $1.4M_{\odot}/1.4M_{\odot}$ BNS
  system.  We have plotted each of the 9 parameters for our
  non-spinning BNS problem as stated in section \ref{waveformSection}.
  Note how, even for parameters with excellent recovery, the peak of
  the 1D Gaussian is displaced from the injected value (in dashed
  red).  This is not due to a systematic bias, but is caused by the
  marginalization of a single dimension from the full 9D posterior
  space.  To better see this effect, compare the 1D PDF for
  $\mathcal{M}_{c}$ and $q$ with the 2D marginalized PDF in
  Figure~\ref{1414masses}}
\end{figure*}

Of the nine parameters in the domain of the waveform, only six are
particularly physically interesting: the masses of the two binaries,
$M_1$ and $M_2$, the orbital inclination, $\iota$, the angular
position on the sky, $\alpha$ and $\delta$, and the luminosity
distance of the source, $D$.  While the coalescence phase $\phi_c$,
the coalescence time $t_c$, and the wave polarization $\psi$ must be
included in any parameter estimation of the waveform, they do not
encode any information of particular astrophysical
interest.\footnote{For systems whose components are non-spinning.  The
  presence of intrinsic angular momentum will, in general, couple to
  the coalescence phase and wave polarization.  \will{But those
    parameters are still physically uninteresting.  They couple to
    interesting parameters in both the spinning and non-spinning
    cases, which is why they must be included.  I would just eliminate
    this footnote.}}

In Figure~\ref{9dPDF}, we provide an example of the nine,
1-dimensional marginalized posterior probability density functions
recovered from a single $1.4M_{\odot}/1.4M_{\odot}$ BNS system.  These
PDFs are typical of the type of results that will be produced by
parameter estimation studies in the advanced detector era.  Notice
that the peak of several parameters, including the chirp mass,
$\chmass$, appears to be displaced from the true values in red.  This
effect is due to the reduction of the 9-dimensional PDF to a series of
marginalized 1-dimensional PDFs.  For instance, the 1-dimensional PDF
for chirp mass is marginalized via
\begin{equation}
p(\chmass | s) = \int _{\thpara \setminus \chmass} p(\thpara |
s)~d(\thpara \setminus \chmass)
\label{mcMarginalization}
\end{equation}
where the notation $\thpara \setminus \chmass$ implies all parameter
of \eqref{parameterspace} except $\chmass$.  Other parameter and
higher-dimensional marginalizations follow a similar convention.  In
practice, the MCMC samples make this integral trivial: since the
samples are distributed according to the posterior,
\eqref{mcMarginalization} can be calculated by simply reading off the
relevant samples in a single parameter, implicitly calculating a
Monte-Carlo integral over all other parameters.

@@

Of the nine variables in our parameter space \eqref{parameterspace},
the mass parameters, $\chmass$ and $q$, or, correspondingly, $M_1$ and
$M_2$ are the most astrophysically interesting.  The ability of
Advanced LIGO/Virgo to construct a population of BNS masses will be
one of the more useful applications of gravitational-wave astronomy.

We find that there is virtually no difference between the mass PDFs of
our different injected signals within each mass bin.  While this may
initially seem disconcerting, it is to be expected.  Recall that all
of the BNS signals were injected with a network SNR of 20 into a
zero-noise detector realization.  Furthermore, note that the mass
parameters are the only two which directly affect the phase of the
TaylorF2 waveform \eqref{phase}.  Therefore, as long as the injected
mass parameters are identical within each mass system, and as long as
the injected SNR is identical, the amount of recoverable information
in each mass parameter should be almost identical, and for a
sufficiently converged MCMC chain, the recovered posterior PDF should
be identical.
